---
title: "oneway"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{varequal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Overview

The `oneway` package provides an integrated workflow for one-factor experimental designs. Its 
functionality covers:

* descriptive statistics;
* *a priori* test (omnibus test)
  - classical ANOVA
  - aligned rank transform ANOVA (`ART-ANOVA`)
  - Kruskal--Wallis test
* *Post hoc* test (Multiple-comparison procedures, MCP): 
  - REGWQ
  - Tukey-HSD
  - Tukey-Kramer
  - Games-Howell
  - Dunn's
* Effect sizes;
* a single adaptive workflow, `pairwise_comparison()`, that selects an appropriate *a priori* + 
  *post hoc* protocol from the observed distributional, variance, and design characteristics.

<br>

# Installation

Install the released version from [CRAN](https://cran.r-project.org/package=oneway):

```{r, eval=FALSE}
install.packages("oneway")
```

or the development version from [GitHub](https://github.com/P10911004-NPUST/oneway):

```{r, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("P10911004-NPUST/oneway")
```

Then load the package:

```{r}
library(oneway)

# For reproducibility
set.seed(123)
```

<br>

# Quick start

The core function of this package is `pairwise_comparison()`, which provides out-of-the-box 
statistical inference for one-way designs. It expects a data frame and a two-sided formula:

```{r}
out <- pairwise_comparison(morphine, tolerance ~ grp)
```

The function return a list which includes 5 components:

- method: the combination of a *a priori* and a *post-hoc* test;
- data: the input data;
- pre_hoc: An ANOVA-like table from the selected *a priori* test;
- post_hoc: pairwise comparison result from the selected *post-hoc* test;
- summary: descriptive statistics for reporting.

The `post_hoc` component is a data frame with at least 15 columns:

| Column name       | Interpretation                                                       |
| :------           | :------------                                                        |
| `x1`              | The name of the group being subtracted                               |
| `x2`              | The name of the group as the subtrahend                              |
| `[x1 - x2]`       | The difference between x1 and x2                                     |
| `Hedges's g`      | Effect size                                                          |
| `Pvalue`          | *P*-value yielded from the standard value                            |
| `Padj`            | adjusted *P*-value according to the `p_adjust_method` argument       |
| `signif`          | Significance symbols. See `oneway:pval2asterisk`                     |
| `diff_CI`         | The (1 - `alpha`) confidence interval of `[x1 - x2]`                 |
| `mu`              | Currently not available                                              |
| `standard_value`  | The statistics of the test, generally is a symbol                    |
| `critical_value`  | The critical value for the statistics                                |
| `StdErr`          | The standard error used to calculate `diff_CI`                       |
| `method`          | The name of the post-hoc method                                      |
| `alternative`     | The direction of testing, either `less`, `greater`, or `two.sided`   |
| `alpha`           | Error tolerance                                                      |


The `summary` component is a data frame with 13 columns:

| Column name   | Interpretation                                                        |
| :---          | :------------                                                         |
| `GROUP`       | The name of each groups from the independent variable                 |
| `CLD`         | Compact letter display, see `oneway::compact_letter_display`          |
| `N`           | Sample size                                                           |
| `AVG`         | Mean                                                                  |
| `SD`          | Sample's standard deviation                                           |
| `MED`         | Median                                                                |
| `MIN`         | Minimum value                                                         |
| `MAX`         | Maximum value                                                         |
| `CI (95%)`    | Confidence interval of including the population's mean                |
| `SKEW (= 0)`  | Skewness, see `normality::skewness`                                   |
| `KURT (= 3)`  | Kurtosis, see `normality::kurtosis`                                   |
| `normality`   | Is the sample follows normal distribution, see `normality::is_normal` |
| `n_outliers`  | Number of possible outliers, see `outlying::Grubbs_test`              |


<h3>If you only require a plug-and-play function to analyze your data, `pairwise_comparison()` 
has you covered. Feel free to skip the rest of this guide.</h3>

<br>

# 1. Interface

Most analysis functions expect a data frame and a two-sided formula:

```r
# Pseudo code, do not run
response ~ group
```

For example:

```{r}
anorexia[c(2, 19, 43, 47, 60, 71), ]
```

The `anorexia` data contain `weight_gain` as the response and `therapy` as the grouping variable:

```{r}
oneway_anova(anorexia, weight_gain ~ therapy, verbose = FALSE)
```

The package also accepts a list of numeric vectors for several functions:

```{r}
group_data <- list(control = rnorm(20, 10, 2),
                   treatment_A = rnorm(20, 12, 2),
                   treatment_B = rnorm(20, 15, 2))

oneway_anova(group_data)
```

<br>

# 2. Describe the data

`describe()` is useful as a first-pass summary.

```{r}
describe(anorexia, weight_gain ~ therapy, rounding = 2)
```

The summary includes:

| Column name   | Interpretation                                                        |
| :---          | :------------                                                         |
| `GROUP`       | The name of each groups from the independent variable                 |
| `CLD`         | Compact letter display, see `oneway::compact_letter_display`          |
| `N`           | Sample size                                                           |
| `AVG`         | Mean                                                                  |
| `SD`          | Sample's standard deviation                                           |
| `MED`         | Median                                                                |
| `MIN`         | Minimum value                                                         |
| `MAX`         | Maximum value                                                         |
| `CI (95%)`    | Confidence interval of including the population's mean                |
| `SKEW (= 0)`  | Skewness, see `normality::skewness`                                   |
| `KURT (= 3)`  | Kurtosis, see `normality::kurtosis`                                   |
| `normality`   | Is the sample follows normal distribution, see `normality::is_normal` |
| `n_outliers`  | Number of possible outliers, see `outlying::Grubbs_test`              |

Note that the `n_outliers` field provides only a **soft suggestion** that approximately *n* 
observations may be outliers. It should therefore be interpreted as a diagnostic flag rather 
than as evidence that those observations should be removed.

<br>

# 3. *a priori* test

## 3.1 Parametric

The classic ANOVA procedure for parametric analysis is `oneway_anova()`.

```{r pre-hoc}
anova_ooo <- oneway_anova(O_O_O, val ~ grp)
anova_ooo
```

The function checks normality and variance homogeneity. With `var_equal = NA` (the default), it
uses `varequal::is_var_equal` to check the variance homogeneity and decide to conduct:

* **Fisher's one-way ANOVA** when equal variances are supported;
* **Welch's ANOVA** when equal variances are not supported.

You can also force the procedure:

```{r}
fisher <- oneway_anova(O_O_O, val ~ grp, var_equal = TRUE)
welch <- oneway_anova(O_X_X, val ~ grp, var_equal = FALSE)
```

Reading the ANOVA table

The Fisher/Welch's ANOVA output includes columns such as:

* `DF`: degrees of freedom;
* `SS`: sum of squares;
* `MS`: mean square;
* `Fvalue`: observed *F* statistic;
* `Fcrit`: critical *F* value;
* `Pvalue`: omnibus *p*-value;
* `signif`: significance label;
* `p_omega2`: partial omega-squared effect size; and
* `method`: Fisher's or Welch's ANOVA.

The automatic selection is convenient for exploratory and routine analyses, but it is still 
important to examine the study design and the diagnostics before interpreting the result.

## 3.2 Nonparametric

### ART-ANOVA

`oneway_art()` implements a one-way aligned-rank-transform analysis.

```{r}
art <- oneway_art(X_O_O, val ~ grp)
art
```

Conceptually, the response is aligned with respect to the group effect, ranked, and then analyzed 
with the ANOVA machinery.

```{r}
attr(art, "data") |> head()
```

ART is useful when the raw response is not adequately compatible with a classical ANOVA but a 
rank-based ANOVA workflow remains appropriate.

### Kruskal--Wallis test

`Kruskal_Wallis_test()` provides the classical nonparametric alternative:

```{r}
kw <- Kruskal_Wallis_test(X_X_O, val ~ grp)
kw
```

<br>

# 4. Post-hoc analysis

`oneway` includes 5 multiple comparison procedures (MCP) for post-hoc analysis. All MCPs produce a 
standardized result identical to the `pairwise_comparison()`.

## 4.1 Parametric

### REGWQ

`REGWQ_test()` implements the Ryan--Einot--Gabriel--Welsch studentized-range procedure. REGWQ is a 
stepwise studentized-range procedure intended primarily for balanced, homoscedastic normal one-way 
designs. The package uses the studentized-range distribution and step-specific significance levels.

```{r}
regwq <- REGWQ_test(O_O_O, val ~ grp)
```

### Tukey HSD

`Tukey_HSD_test()` provides the classical Tukey honestly significant difference procedure. The 
function is intended primarily for balanced designs with approximately normal responses and 
homogeneous variances. When the design is unbalanced, use Tukey--Kramer instead.

```{r}
tukey <- Tukey_HSD_test(O_O_O, val ~ grp)
```

### Tukey--Kramer

`Tukey_Kramer_test()` extends the Tukey's comparison to unequal sample sizes while retaining the 
equal-variance assumption:

```{r}
tukey_kramer <- Tukey_Kramer_test(O_O_X, val ~ grp)
```


### Games--Howell

`Games_Howell_test()` is intended for normal data with unequal variances and/or unequal sample 
sizes. Each comparison uses a Welch--Satterthwaite-type degrees of freedom and the 
studentized-range distribution. This makes Games--Howell the natural choice among the package's 
parametric post-hoc procedures when homoscedasticity assumption is violated.

```{r}
games_howell <- Games_Howell_test(O_X_X, val ~ grp)
```

## 4.2 Nonparametric

### Dunn's test

`Dunn_test()` is the nonparametric multiple-comparison procedure associated with Kruskal--Wallis 
analysis.

```{r}
dunn <- Dunn_test(X_X_O, val ~ grp, p_adjust_method = "holm")
```

Dunn's procedure compares mean ranks. Because many pairwise comparisons are performed, the function 
supports `stats::p.adjust()` methods, with `"holm"` as the default:

```{r}
dunn_bh <- Dunn_test(X_X_O, val ~ grp, p_adjust_method = "BH")
```

The available choices are those accepted by `stats::p.adjust()`.

<br>

# 5. `pairwise_comparison` workflow

The package's main automatic workflow is `pairwise_comparison()`.

```{r}
adaptive <- pairwise_comparison(O_O_O, val ~ grp)
```

The decision logic is:

| Data characteristics                             | Omnibus / pre-hoc | Post-hoc          |
| ---                                              | ---               | ---               |
| Normal, equal variance, balanced                 | Fisher's ANOVA    | REGWQ             |
| Normal, equal variance, unbalanced               | Fisher's ANOVA    | Tukey--Kramer     |
| Normal, unequal variance                         | Welch's ANOVA     | Games--Howell     |
| Non-normal, ART rank response becomes suitable   | ART-ANOVA         | similar to normal |
| Non-normal, ART rank response remains unsuitable | Kruskal--Wallis   | Dunn              |

The adaptive workflow can therefore return 7 complete protocols:

1. one-way ANOVA + REGWQ;
2. one-way ANOVA + Tukey--Kramer;
3. one-way ANOVA + Games--Howell;
4. ART-ANOVA + [REGWQ on y''];
5. ART-ANOVA + [Tukey--Kramer on y''];
6. ART-ANOVA + [Games--Howell on y'']; or
7. Kruskal--Wallis + Dunn.

The Fisher's and Welch's ANOVA procedures are conducted on the residuals rather than the raw 
response.

<br>

# 6. Compact letter displays

A compact letter display is a presentation device:

* groups **sharing a letter** are **NOT** significantly different;
* groups without a common letter are significantly different.

`compact_letter_display()` can be used independently. For example:

```{r}
dunn <- Dunn_test(X_X_O, val ~ grp, verbose = FALSE)

summary_tab <- row_arrange(dunn$summary, cols = "MED")

padj_name <- "Padj (holm)"

cld <- compact_letter_display(x1 = dunn$post_hoc$x1,
                              x2 = dunn$post_hoc$x2,
                              pvalues = dunn$post_hoc[[padj_name]],
                              grp_names = summary_tab$GROUP,
                              centers = summary_tab$MED)

cld
```

The `centers` argument controls the order used to assign letters. For parametric procedures it is 
typically natural to use the mean; for rank-based procedures, the median is more appropriate.

The symbols can also be customized:

```{r}
compact_letter_display(
  x1 = dunn$post_hoc$x1,
  x2 = dunn$post_hoc$x2,
  pvalues = dunn$post_hoc[[padj_name]],
  grp_names = summary_tab$GROUP,
  centers = summary_tab$MED,
  display_letters = as.character(1:6),
  display_null_letter = "_"
)
```

For two-samples comparison, use `pval2asterisk()` to convert numeric p-values into user-defined
significance symbols.

```{r}
p <- c(0.20, 0.04, 0.008, 0.0005, 1e-6)

pval2asterisk(p)
```

The default labels are:

| Range                | Label   |
| :---------           | :----:  |
| p > 0.055            | `ns`    |
| 0.05 < p &le; 0.055  | `.`     |
| 0.01 < p &le; 0.05   | `*`     |
| 0.001 < p &le; 0.01  | `**`    |
| p &le; 0.001         | `***`   |

Custom thresholds and symbols are supported. The function expects thresholds in descending order:

```{r}
pval2asterisk(p,
              break_points = c(0.05, 0.01, 0),
              symbols = c("NS", "S", "HS"))
```

# 7. Effect sizes

This package provides quick inference to the effect size using only 2 of the many estimators.
For comprehensive effect size analysis, please refer to the paper of
[Lakens, 2013](https://doi.org/10.3389/fpsyg.2013.00863) and the
[effectsize](https://cran.r-project.org/package=effectsize) CRAN package.

### Hedges' *g*

`Hedges_g_s()` applies a small-sample correction to Cohen's *d*:

```{r}
g1 <- subset(O_O_O, grp == "G1")$val
g2 <- subset(O_O_O, grp == "G2")$val
Hedges_g_s(g1, g2)
```

Conventional benchmarks:

| Hedges' *g* | Effect | Practical meaning                                        |
| ---         | ---    | ---------                                                |
| 0.2         | Small  | The difference is noticeable but modest.                 |
| 0.5         | Medium | The difference is clearly visible.                       |
| 0.8         | Large  | The difference is substantial and practically important. |

### Partial-&omega;<sup>2</sup>

The classical ANOVA implementation reports partial omega-squared as `p_omega2`:

```{r}
fisher[, c("method", "Pvalue", "p_omega2")]
```

<br>

# 8. Datasets

The package includes 6 simulated datasets designed specifically for different analysis conditions:

| Dataset | Distribution | Variance      | Balance    |
| :---:   | :---:        | :---:         | :---:      |
| `O_O_O` | Normal       | Homogeneous   | Balanced   |
| `O_O_X` | Normal       | Homogeneous   | Unbalanced |
| `O_X_X` | Normal       | Heterogeneous | Unbalanced |
| `X_O_O` | Non-normal   | Homogeneous   | Balanced   |
| `X_X_O` | Non-normal   | Heterogeneous | Balanced   |
| `X_X_X` | Non-normal   | Heterogeneous | Unbalanced |

Each has six groups (`G1`--`G6`) and uses the common `grp` / `val` column names.

Additional datasets are:

* `anorexia`: weight gain following three therapies;
* `morphine`: morphine tolerance across five treatment groups; and
* `plasma_etching`: etch rate across plasma power levels.

<br>

# 9. Interpretation notes

### Statistical significance is not the same as practical importance

A small *p*-value indicates evidence against the null hypothesis under the specified model. Effect 
sizes and confidence intervals provide complementary information about magnitude and precision.

### CLDs are a summary, not the primary inferential result

Use the pairwise table to inspect the actual differences, adjusted *p*-values, confidence intervals, 
and effect sizes. Treat the CLD as a compact visualization of those decisions.

### Do not mechanically transform data merely to obtain a desired p-value

The package provides rank-based alternatives so that a scientifically defensible analysis does not 
depend on forcing the response into a particular distributional shape.

# Summary

A practical way to use `oneway` is:

1. Confirm that the design is genuinely one-way: 
   one independent grouping variable and one quantitative response variable.
2. Inspect sample sizes, descriptive statistics, possible outliers, and distributional shape.
3. Use `oneway_anova()` when a parametric omnibus analysis is appropriate.
4. Use `oneway_art()` when an aligned-rank-transform analysis is justified.
5. Use `Kruskal_Wallis_test()` when a rank-based omnibus test is more appropriate.
6. Choose a post-hoc procedure that matches the assumptions and balance of the design.
7. Report the estimated pairwise differences, confidence intervals, adjusted *p*-values, and 
   effect sizes (not only significance stars or CLDs).
8. Use `pairwise_comparison()` when you want to formalize this decision process automatically.

The package's automatic workflow is particularly useful when you analyze many similar one-way 
experiments and want a consistent rule for selecting the analysis protocol.

<br>

# References

Dinno, A. (2015). 
Nonparametric pairwise multiple comparisons in independent groups using Dunn’s test. 
*The Stata Journal: Promoting Communications on Statistics and Stata*, 15(1), 292–300. <https://doi.org/10.1177/1536867X1501500117>

Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). 
An aligned rank transform procedure for multifactor contrast tests. 
*Proceedings of the 34th Annual ACM Symposium on User Interface Software and Technology*, 754–768. <https://doi.org/10.1145/3472749.3474784>

Hollander, M., Wolfe, D. A., & Chicken, E. (2014). 
*Nonparametric statistical methods* (3rd ed.). Wiley.

Howell, D. C. (2010). *Statistical methods for psychology* (7th ed.). Cengage Learning.

Howell, D. C. (2013). *Statistical methods for psychology* (8th ed.). Cengage Learning.

Kruskal, W. H., & Wallis, W. A. (1952). 
Use of ranks in one-criterion variance analysis. 
*Journal of the American Statistical Association*, 47(260), 583–621. 
<https://doi.org/10.1080/01621459.1952.10483441>

Lakens, D. (2013). 
Calculating and reporting effect sizes to facilitate cumulative science:
A practical primer for t-tests and ANOVAs. 
*Frontiers in Psychology*, 4(863). 
<https://doi.org/10.3389/fpsyg.2013.00863>

Montgomery, D. C. (2017). *Design and analysis of experiments* (9th ed.). John Wiley & Sons.

Piepho, H.-P. (2004). 
An algorithm for a letter-based representation of all-pairwise comparisons. 
*Journal of Computational and Graphical Statistics*, 13(2), 456–466. 
<https://doi.org/10.1198/1061860043515>

Piepho, H.-P. (2018). 
Letters in mean comparisons: What they do and don’t mean. 
*Agronomy Journal*, 110(2), 431–434. 
<https://doi.org/10.2134/agronj2017.10.0580>

Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). 
The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. 
*Proceedings of the SIGCHI Conference on Human Factors in Computing Systems*, 143–146. <https://doi.org/10.1145/1978942.1978963>

Zar, J. H. (2014). *Biostatistical analysis* (5th ed.). Pearson.

